
---
title: "Kendrick Lamar vs. Drake: Data Analysis Report"
author: "Aamodit Acharya"
date: "`r Sys.Date()`"
output: html_document
---

## Introduction

This report explores the relative popularity of Kendrick Lamar and Drake by analyzing their songs' play counts on YouTube Music. We visualize their individual performance using **scatter plots** and **boxplots** to identify patterns. Additionally, we apply **k-Nearest Neighbors (kNN) classification** to understand whether song play counts can predict the artist.

---

## 1. Scatter Plot: Kendrick Lamar vs. Drake

We start with a **scatter plot** comparing Kendrick Lamar and Drake's songs, using the number of plays (in millions) on YouTube Music. The plot displays each artist’s song with a jittered index to avoid overlapping points.

### Scatter Plot: Number of Plays vs. Song Index

```{r scatter-plot, echo=FALSE, fig.align='center', fig.cap="Scatter Plot of Song Plays for Kendrick Lamar and Drake"}
# Load necessary libraries
if (!require(scales)) install.packages("scales", dependencies = TRUE)
library(scales)

# Reindex Kendrick's and Drake's songs to start from 1
kendrick_plays$song_index <- seq_along(kendrick_plays$Title)
drake_plays$song_index <- seq_along(drake_plays$Title)

# Plot Kendrick's songs
plot(
  jitter(kendrick_plays$song_index, factor = 3),
  kendrick_plays$Plays / 1e6,
  col = alpha("blue", 0.5),
  pch = 19,
  xlab = "Song Index",
  ylab = "Artist Plays in millions",
  main = "Number of Plays: Drake vs Kendrick",
  xlim = c(0, max(nrow(kendrick_plays), nrow(drake_plays))),
  ylim = c(0, 2200)
)

# Add Drake's songs to the plot
points(
  jitter(drake_plays$song_index, factor = 3),
  drake_plays$Plays / 1e6,
  col = alpha("red", 0.5),
  pch = 19
)

# Add a legend
legend(
  "topright",
  legend = c("Kendrick Lamar", "Drake"),
  col = alpha(c("blue", "red"), 0.5),
  pch = 19
)
```

### Discussion

- The scatter plot shows that **most of Drake’s songs** accumulate towards the right side of the plot (higher song index), reflecting his larger catalog.
- **Both artists** have a few songs with extremely high plays, shown by points reaching the top of the Y-axis.
- Kendrick’s songs are more clustered towards the left, suggesting a smaller number of high-performing tracks compared to Drake.

---

## 2. Boxplot Comparison: Kendrick Lamar vs. Drake

Next, we generate **side-by-side boxplots** to compare the distribution of song plays for each artist.

### Boxplot: Distribution of Plays for Kendrick and Drake

```{r boxplot, echo=FALSE, fig.align='center', fig.cap="Boxplots of Song Plays for Kendrick Lamar and Drake"}
par(mfrow = c(1, 2))

# Boxplot for Kendrick's songs
boxplot(
  kendrick_plays$Plays / 1e6,
  xlab = "Songs with Kendrick",
  ylab = "Plays in millions",
  main = "Number of Plays for Kendrick"
)

# Boxplot for Drake's songs
boxplot(
  drake_plays$Plays / 1e6,
  xlab = "Songs with Drake",
  ylab = "Plays in millions",
  main = "Number of Plays for Drake"
)

# Reset plotting layout
par(mfrow = c(1, 1))
```

### Discussion

- The **median play count** for Drake’s songs is noticeably higher than that of Kendrick Lamar.
- Drake’s songs also show **higher variability**, with some songs achieving **over 2 billion plays**.
- Kendrick Lamar’s distribution shows more consistency, though he has a few high outliers.

---

## 3. kNN Classification: Predicting the Artist Based on Plays

We apply the **k-Nearest Neighbors (kNN) classification** to predict whether a song belongs to Kendrick Lamar or Drake based on the number of plays it has received.

### Confusion Matrix

```{r confusion-matrix, echo=FALSE}
# Define the kNN and helper functions here (as provided earlier)

# Prepare the data
x <- kvd_data$Plays_mil
y <- kvd_data$Artist_Group_Bin

# Run kNN with k = 5
results <- knn(x, y, k = 5)

# Confusion matrix calculations
TP <- sum(y == 1 & results == 1)  # True Positive
FN <- sum(y == 1 & results == 0)  # False Negative
FP <- sum(y == 0 & results == 1)  # False Positive
TN <- sum(y == 0 & results == 0)  # True Negative

# Create confusion matrix
confusion_matrix <- matrix(c(TP, FN, FP, TN), nrow = 2, byrow = TRUE)
colnames(confusion_matrix) <- c("Predicted Kendrick", "Predicted Drake")
rownames(confusion_matrix) <- c("Actual Kendrick", "Actual Drake")
print(confusion_matrix)
```

### Discussion

- The confusion matrix shows that the model performs well, correctly classifying most of the songs.
- However, the model tends to **misclassify high-play Kendrick songs as Drake's**, indicating that the model may be biased toward Drake’s catalog due to the higher average plays.

---

## 4. Accuracy vs. k Plot

We examine the effect of varying the number of neighbors (**k**) in the kNN classifier and plot the **accuracy vs. k**.

### Plot: Accuracy vs. k

```{r accuracy-plot, echo=FALSE, fig.align='center', fig.cap="Plot of Accuracy vs. k Values"}
accuracy_list <- list()
k_values <- 2:200

# Calculate accuracy for each k value
for (k in k_values) {
  results <- knn(x, y, k)
  TP <- sum(y == 1 & results == 1)
  FN <- sum(y == 1 & results == 0)
  FP <- sum(y == 0 & results == 1)
  TN <- sum(y == 0 & results == 0)
  accuracy <- (TP + TN) / (TP + TN + FP + FN)
  accuracy_list <- append(accuracy_list, accuracy)
}

# Plot accuracy vs. k
plot(
  k_values, unlist(accuracy_list),
  type = "b", pch = 19,
  main = "K Value Accuracy",
  xlab = "Number of Neighbours (k values)",
  ylab = "Accuracy"
)
```

### Discussion

- The plot shows that **accuracy peaks around k = 10 to 30**, after which it gradually declines.
- With **higher values of k**, the classifier becomes less effective, likely due to over-generalization.
- **Lower values of k** lead to overfitting, where the model memorizes individual points but fails to generalize.

---

## Conclusion

This report provides a comprehensive analysis of the popularity of Kendrick Lamar and Drake through visualizations and machine learning models. We observed:
- Drake tends to have more high-play songs overall, while Kendrick has fewer but still significant hits.
- The kNN classification can reasonably predict the artist based on song plays, but higher play counts sometimes skew the results toward Drake.

---

## References

- Dataset: YouTube Music statistics as of May 2024